---
title: "Excess mortality & voting patterns in CH"
subtitle: "Mortality data preparation (*Wohnort* based)"
pagetitle: "Excess & voting: mortality data"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(tidyverse, magrittr, 
       janitor, scales, lubridate, 
       DT, skimr)

import::from("sjmisc", "frq")
```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)
```

<!-- ----------------------------------------------------- -->

# Source 

Custom file delivered by BfS covering 2021 period and standardizing municipalities to Jul 2021.  

Using **Wohnort** (*the municipality where the person lived*) as unit of analysis.   

```{r}
w_deaths_2015_2021 <- read_csv("data-raw/BfS-closed/monthly_deaths/13OCT21_Lieferung_Panczak_Wohnort.zip", 
                               col_types = cols(year = col_integer(), 
                                                month = col_integer(), 
                                                COUNT = col_integer()),
                               locale = readr::locale(encoding = "latin1")) %>% 
  mutate(sex = if_else(sex == "F", "Female", "Male")) %>% 
  select(-canton, -community, -nationality) %>% 
  rename(GMDNAME = wohnort,
         canton = wohnkanton) %>% 
  # hyphen in name >> doesnt link to spatial data
  mutate(GMDNAME = if_else(GMDNAME == "La Grande-Béroche", "La Grande Béroche", GMDNAME)) %>% 
  # name change? >> resolved with Cordula
  mutate(GMDNAME = if_else(GMDNAME == "Klosters-Serneus", "Klosters", GMDNAME)) 
```

# Municipality updates 

To reach the state of `2022-01-01`.  

```{r}
w_deaths_2015_2021 %<>% 
  # Essertes merge
  mutate(GMDNAME = if_else(GMDNAME == "Essertes", "Oron", GMDNAME)) %>% 
  # Zurzach
  mutate(GMDNAME = if_else(GMDNAME == "Bad Zurzach", "Zurzach", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Baldingen", "Zurzach", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Böbikon", "Zurzach", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Kaiserstuhl", "Zurzach", GMDNAME)) %>%
  mutate(GMDNAME = if_else(GMDNAME == "Rekingen (AG)", "Zurzach", GMDNAME)) %>%
  mutate(GMDNAME = if_else(GMDNAME == "Rietheim", "Zurzach", GMDNAME)) %>%
  mutate(GMDNAME = if_else(GMDNAME == "Rümikon", "Zurzach", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Wislikofen", "Zurzach", GMDNAME)) %>% 
  # Böztal merge
  mutate(GMDNAME = if_else(GMDNAME == "Bözen", "Böztal", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Effingen", "Böztal", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Elfingen", "Böztal", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Hornussen", "Böztal", GMDNAME)) %>% 
  # Blonay - Saint-Légier merge
  mutate(GMDNAME = if_else(GMDNAME == "Blonay", "Blonay - Saint-Légier", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Saint-Légier-La Chiésaz", "Blonay - Saint-Légier", GMDNAME)) %>% 
  # Murten merge
  mutate(GMDNAME = if_else(GMDNAME == "Galmiz", "Murten", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Gempenach", "Murten", GMDNAME)) %>% 
  mutate(GMDNAME = if_else(GMDNAME == "Clavaleyres", "Murten", GMDNAME))
```

# Monthly aggregation

```{r}
w_deaths_2015_2021 %<>% 
  group_by(year, month, canton, GMDNAME, agegroup, sex) %>% 
  summarise(deaths = sum(COUNT)) %>% 
  ungroup()
```

## Description of data  

```{r echo=FALSE}
report::report(w_deaths_2015_2021)
```

```{r echo=FALSE}
skim(w_deaths_2015_2021)
```

```{r eval=FALSE, include=FALSE}
# Deaths across variables:  
w_deaths_2015_2021 %>%
  mutate(month = factor(month)) %>% 
  uncount(deaths) %>% 
  select(-GMDNAME, -canton) %>% 
  gtsummary::tbl_summary(by = year)
```

```{r eval=FALSE, include=FALSE}
# Canton x year distribution:  
w_deaths_2015_2021 %>%
  uncount(deaths) %>% 
  select(year, canton) %>% 
  gtsummary::tbl_summary(by = year)
```

`r number(sum(w_deaths_2015_2021$deaths), big.mark = ",")` deaths by year:  

```{r echo=FALSE}
w_deaths_2015_2021 %>%
  group_by(year) %>% 
  summarise(Deaths = sum(deaths)) %>% 
  mutate(Deaths = as.character(number(Deaths, big.mark = ","))) 
```

# Codes of geographical areas

```{r include=FALSE}
raum <- read_rds("data/BfS/raum.Rds")
```

```{r include=FALSE}
# Codes with no link - has to be empty 

no_link <- w_deaths_2015_2021 %>% 
  select(GMDNAME) %>% 
  distinct()  %>% 
  anti_join(raum)

stopifnot(nrow(no_link) == 0)

rm(no_link)
```

There were `r number(nrow(raum), big.mark = ",")` municipalities on the `2022-01-01` in CH.  

```{r echo=FALSE}
no_cases <- raum %>% 
  anti_join(
    w_deaths_2015_2021 %>% 
      select(GMDNAME) %>% 
      distinct() 
  )
```

`r number(nrow(no_cases))` had no deaths recorded during the study period.   

```{r echo=FALSE}
no_cases %>% 
  select(GMDNAME, KTNAME) %>% 
  arrange(GMDNAME)

rm(no_cases)
```

# Age groups

Original age groups:  

```{r echo=FALSE}
frq(w_deaths_2015_2021, agegroup)
```

Were simplified to `<40`, `40-59` , `60-69`, `70-79` & `80+`

```{r}
w_deaths_2015_2021 %<>% 
  mutate(
    age = case_when(
      agegroup == "[0-10)" ~   "<40",
      agegroup == "[10-20)" ~  "<40",
      agegroup == "[20-30)" ~  "<40",
      agegroup == "[30-40)" ~  "<40",
      
      agegroup == "[40-50)" ~  "40-49",
      agegroup == "[50-60)" ~  "50-59",
      agegroup == "[60-70)" ~  "60-69",
      agegroup == "[70-80)" ~  "70-79",
      
      agegroup == "[80-90)" ~  "80+",
      agegroup == "[90-116)" ~ "80+" ,
      TRUE ~                    agegroup)
  ) %>% 
  select(-agegroup) %>% 
  group_by(GMDNAME, year, month, canton, sex, age) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup()
```

```{r echo=FALSE}
frq(w_deaths_2015_2021, age)
```

# Distribution of death counts

Obviously highly skewed.  
**Note:** *No zeroes here yet! No pop denominator either!*  

```{r echo=FALSE}
w_deaths_2015_2021 %>% 
  ggplot(aes(deaths)) + 
  geom_histogram(binwidth = 1)
```

## Filling with zeroes

```{r eval=FALSE, include=FALSE}
raum %>% 
  group_by(GMDNAME) %>% 
  filter(n() > 1)
```

All possible combinations of `year, month, canton, municipality, age & sex` are used and then empty strata are filled with zeroes.  

This expansion also includes municipalities with zero deaths described above.  

```{r}
w_deaths_2015_2021_exp <- 
  raum %>% 
  select(GMDNAME) %>% 
  distinct() %>% 
  left_join(w_deaths_2015_2021) %>% 
  select(-deaths, -canton) %>% 
  # avoiding NAs in three communities without cases
  mutate(
    year = if_else(is.na(year), as.integer(2015), year),
    month = if_else(is.na(month), as.integer(1), month),
    sex = if_else(is.na(sex), "Male", sex),
    age = if_else(is.na(age), "<40", age)
  ) %>% 
  expand(GMDNAME, year, month, sex, age) %>%
  left_join(w_deaths_2015_2021) %>% 
  replace_na(list(deaths = 0)) %>% 
  mutate(date = ymd(paste(year, month, "1", sep = " "))) %>% 
  relocate(date, .after = month) %>% 
  left_join(raum) %>% 
  select(-canton) %>% 
  relocate(GMDNR, .before = GMDNAME) %>% 
  relocate(KTNAME, .after = GMDNAME) %>% 
  arrange(GMDNR, year, month, age, sex) %>% 
  filter( !(year == 2021 & month > 6))
```

```{r eval=FALSE, include=FALSE}
length(unique(raum$GMDNAME))
length(unique(w_deaths_2015_2021_exp$GMDNAME))
```

That obviously increases the size of the dataset from `r number(nrow(w_deaths_2015_2021), big.mark = ",")` to `r number(nrow(w_deaths_2015_2021_exp), big.mark = ",")` (*ie.* 2,163 municipalities each with 2,016 data points * 7 years * 12 months * 6 age groups * 2 sexes).  

## Zero cells 

Needless to say that creates large amount of zero cells:

```{r echo=FALSE}
w_deaths_2015_2021 %>% 
  ggplot(aes(deaths)) + 
  geom_histogram(binwidth = 1)
```

```{r echo=FALSE}
frq(w_deaths_2015_2021_exp, deaths == 0)
```

Situation only marginally improves with exclusion of `<40` age group stratum:  

```{r echo=FALSE}
w_deaths_2015_2021_exp %>% 
  filter(age != "<40") %>% 
  frq(deaths == 0)
```

# Examples of time series

## Municipalities Bern & Meienried:  

```{r echo=FALSE}
w_deaths_2015_2021_exp %>% 
  filter(GMDNAME %in% c("Bern", "Meienried")) %>% 
  group_by(GMDNAME, date) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = deaths, group = GMDNAME, color = GMDNAME)) + 
  geom_line() + geom_point() +
  theme_minimal()
```

## All time series

```{r echo=FALSE}
w_deaths_2015_2021_exp %>% 
  group_by(GMDNAME, date) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = deaths, group = GMDNAME)) + 
  geom_line(colour = "darkorchid", alpha = 0.25) +
  theme_minimal()
```

# Yearly aggregation

Excluding 2021.  

```{r}
w_deaths_2015_2020_year <- w_deaths_2015_2021_exp %>% 
  filter(year < 2021) %>% 
  group_by(year, KTNAME, GMDNR, GMDNAME, 
           ARGRNR, ARGRNAME, 
           ARNR, ARNAME, border, 
           sex, age) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup()
```

## Miscelaneous ID variables for INLA

```{r}
gg <- read_rds("data/BfS/gg.Rds") %>% 
  sf::st_drop_geometry() %>% 
  select(GMDNR, id_space)

w_deaths_2015_2020_year %<>% 
  left_join(gg) %>% 
  mutate(id_year = year - 2014,
         id_age = as.integer(as.factor(age)),
         id_2015 = if_else(year == 2015, 1, 0), 
         observed = deaths, 
         deaths = if_else(year == 2020, NA_integer_, deaths)) %>% 
  relocate(observed, .after = deaths) %>% 
  relocate(id_space, .after = GMDNR) %>% 
  relocate(id_age, .after = age) %>% 
  relocate(id_year, .after = year) %>% 
  relocate(id_2015, .after = id_year) 
```

## Description of data  

```{r echo=FALSE}
w_deaths_2015_2020_year %>% 
  select(-starts_with("id_")) %>% 
  report::report()
```

```{r echo=FALSE}
skim(w_deaths_2015_2020_year)
```

```{r include=FALSE}
write_rds(w_deaths_2015_2020_year, "data/BfS-closed/monthly_deaths/w_deaths_2015_2020_year.Rds")
```

That obviously decreases the size of the dataset from `r number(nrow(w_deaths_2015_2021_exp), big.mark = ",")` to `r number(nrow(w_deaths_2015_2020_year), big.mark = ",")` (*ie.* 2,163 municipalities each with 2,016 data points * 6 years * 6 age groups * 2 sexes).  

## Zero cells 

Less but still large amount of zero cells:

```{r echo=FALSE}
w_deaths_2015_2020_year %>% 
  ggplot(aes(deaths)) + 
  geom_histogram(binwidth = 1)
```

```{r echo=FALSE}
frq(w_deaths_2015_2020_year, deaths == 0)
```

Situation only marginally improves with exclusion of `<40` age group stratum:  

```{r echo=FALSE}
w_deaths_2015_2020_year %>% 
  filter(age != "<40") %>% 
  frq(deaths == 0)
```

## Examples of Bern & Meienried:  

```{r echo=FALSE}
w_deaths_2015_2020_year %>% 
  filter(GMDNAME %in% c("Bern", "Meienried")) %>% 
  group_by(GMDNAME, year) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = year, y = deaths, group = GMDNAME, color = GMDNAME)) + 
  geom_line() + geom_point() +
  theme_minimal()
```

# Groups for stratified modelling 

```{r}
groups = as_tibble(expand_grid(age = c("<40", "40-49", "50-59", 
                                       "60-69", "70-79", "80+"),
                               sex = c("Female", "Male")))
```

```{r include=FALSE}
write_rds(groups, "data/groups.Rds")
```