---
title: "2020 excess mortality & voting patterns in CH"
subtitle: "Redistributed cantonal deaths"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(tidyverse, magrittr, skimr, scales,
       hrbrthemes, 
       sf, tmap,
       osfr)

import::from("sjmisc", "frq")
```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)
```

<!-- ----------------------------------------------------- -->

# Data 

## Canton 

Expected deaths by age X sex X canton in 2020 from [Riou *et al.*](https://www.medrxiv.org/content/10.1101/2022.08.05.22278458v1), now published as [Riou *et al.*](https://pubmed.ncbi.nlm.nih.gov/36609356/). We correct the previous mistake by taking the expected deaths by age, sex and canton for the whole of 2020, not just from February 24th, 2020 to December 31st, 2020.

```{r include=FALSE}
exp_deaths_2020_kt = read_rds("data/canton/exp_deaths_2020_kt_v2.rds") %>% select(-year)

canton_abbreviations <- c(
  "AG" = "AG", "Appenzell Ausserrhoden" = "AR", "Appenzell Innerrhoden" = "AI", 
  "Basel-Landschaft" = "BL", "Basel-Stadt" = "BS", "Bern / Berne" = "BE", 
  "Fribourg / Freiburg" = "FR", "Genève" = "GE", "Glarus" = "GL", 
  "Graubünden / Grigioni / Grischun" = "GR", "Jura" = "JU", "Luzern" = "LU", 
  "Neuchâtel" = "NE", "Nidwalden" = "NW", "Obwalden" = "OW", 
  "Schaffhausen" = "SH", "Schwyz" = "SZ", "Solothurn" = "SO", 
  "St. Gallen" = "SG", "Thurgau" = "TG", "Ticino" = "TI", "Uri" = "UR", 
  "Valais / Wallis" = "VS", "Vaud" = "VD", "Zug" = "ZG", "Zürich" = "ZH"
)
exp_deaths_2020_kt = exp_deaths_2020_kt %>%
  mutate(canton = recode(canton, !!!canton_abbreviations))

```

Example of one iteration  

```{r eval=FALSE, include=FALSE}
exp_deaths_2020_kt %>% 
  filter(it == 1) %>% 
  filter(age_group != "<40") %>% 
  summarise(exp=sum(exp_deaths),obs=sum(observed2))
```

The median total expected deaths for the entire year is now 64,406 (instead of 55,676 before the correction).

```{r}
exp_deaths_2020_kt %>% 
  filter(age_group != "<40") %>% 
  group_by(it) %>% 
  summarise(exp=sum(exp_deaths), .group="drop") %>% 
  summarise(`Expected (median)`=round(median(exp)),
            `Expected (lower bound)`=round(quantile(exp,0.025)),
            `Expected (upper bound)`=round(quantile(exp,0.975)),.groups="drop" )
```


## Municipality 

Load municipality level data.

```{r}
x = read_rds("results/exp_deaths_2020_year.Rds")
x2 = x %>%  
  filter(it==1) %>% 
  group_by(canton, GMDNR, age_group, sex) %>% 
  summarise(observed = sum(munici_observed ),
            pop_mid_poi = sum(munici_pop),
            .groups="drop")
```

```{r echo=FALSE}
x %>% 
  skim()
```

## Spatial 

```{r}
gg = read_rds("data/BfS/gg.Rds")
```

## Downscale function 

@jriou method to downscale cantonal deaths to municipality level.  

```{r}
source("R/downscale_year.R")
```

<!-- ----------------------------------------------------- -->

# Downscale 

```{r}
exp_deaths_2020_year = downscale_year(exp_deaths_2020_kt, x2) %>% 
  rename(
    munici_pop = pop_mid_poi,
    munici_observed = observed,
    cant_exp_deaths = exp_deaths,
    cant_observed = observed2) 
```

Again, result contains iterations of diff results  

```{r}
max(exp_deaths_2020_year$it)
```

Example of one iteration  

```{r echo=FALSE}
exp_deaths_2020_year %>% 
  filter(it == 1) %>% 
  select(obs=5,exp=11) %>%
  summarize(exp=sum(exp),obs=sum(obs)) 
```

Over all iterations:

```{r}
exp_deaths_2020_year %>% 
  group_by(it) %>% 
  summarise(exp=sum(munici_exp_deaths), obs=sum(munici_observed),.group="drop") %>% 
  summarise(Observed=median(obs),
            `Expected (median)`=round(median(exp)),
            `Expected (lower bound)`=round(quantile(exp,0.025)),
            `Expected (upper bound)`=round(quantile(exp,0.975)),.groups="drop" )
```

```{r include=FALSE}
rm(exp_deaths_2020_kt, w_deaths_2020_year_fin, downscale_year); gc()
```

<!-- ----------------------------------------------------- -->

# Preps

## Aggregates 

Iterations collapsed for each municipality strata.  

```{r}
exp_deaths_2020_year_agg = exp_deaths_2020_year %>% 
  group_by(canton, GMDNR, age_group, sex) %>% 
  summarise(
    # observed
    munici_observed = first(munici_observed),
    munici_pop = first(munici_pop),
    # modelled x1000
    # expected
    munici_expected_lo = quantile(munici_exp_deaths, 0.025),
    munici_expected_med = quantile(munici_exp_deaths, 0.5),
    munici_expected_up = quantile(munici_exp_deaths, 0.975),
    # excess
    munici_excess_lo = quantile(munici_excess, 0.025),
    munici_excess_med = quantile(munici_excess, 0.5),
    munici_excess_up = quantile(munici_excess, 0.975)
  ) %>% 
  ungroup()
```

Even simpler aggregation to municipality level, ignoring age & sex strata.  

```{r}
exp_deaths_2020_year_gem = exp_deaths_2020_year_agg %>% 
  group_by(canton, GMDNR) %>% 
  summarise(
    # community level totals
    munici_pop = sum(munici_pop),
    munici_observed = sum(munici_observed),
    munici_expected_med = sum(munici_expected_med),
    munici_excess_med = sum(munici_excess_med)
  ) %>% 
  ungroup() %>% 
  # ratio and per pop
  mutate(munici_excess_rat = munici_excess_med / munici_expected_med,
         munici_excess_pop =  (munici_excess_med / munici_pop) * 1000) 
```

Special situation arises in this aggregation when calculating a *ratio of excess deaths to observed* deaths. There are communities where expected number of deaths is zero. That leads to ratio being `Inf` in cases where excess is > 0 and to `NaN` where excess is (also) == 0.  

```{r echo=FALSE}
# frq(exp_deaths_2020_year_gem, munici_expected_med == 0 & munici_excess_med != 0)
# frq(exp_deaths_2020_year_gem, munici_expected_med == 0 & munici_excess_med == 0)
frq(exp_deaths_2020_year_gem, munici_expected_med == 0)
```

These communities" ratios were recoded to `NA`.  

```{r include=FALSE}
exp_deaths_2020_year_gem %<>% 
  mutate(munici_excess_rat = ifelse(munici_expected_med == 0, NA, munici_excess_rat))
```

## Bring covariates 

```{r}
covs1 = read_rds("results/exp_deaths_2020_year.Rds")

covs1 = covs1 %>% 
  dplyr::arrange(it,canton,GMDNR,age_group,sex) %>% 
  dplyr::select(GMDNAME,id_space,15:28)
exp_deaths_2020_year_v2 = exp_deaths_2020_year %>% 
  dplyr::arrange(it,canton,GMDNR,age_group,sex) %>% 
  dplyr::bind_cols(covs1)

covs2 = read_rds("results/exp_deaths_2020_year_agg.Rds")
covs2 = covs2 %>% 
  dplyr::arrange(canton,GMDNR,age_group,sex) %>% 
  dplyr::select(canton,GMDNR,GMDNAME,age_group,sex,14:26)
exp_deaths_2020_year_agg_v2 = exp_deaths_2020_year_agg %>% 
  dplyr::arrange(canton,GMDNR,age_group,sex) %>% 
  dplyr::left_join(covs2)

covs3 = read_rds("results/exp_deaths_2020_year_gem.Rds")
covs3 = covs3 %>% 
  dplyr::arrange(canton,GMDNR) %>% 
  dplyr::select(canton,GMDNR,GMDNAME,10:22)
exp_deaths_2020_year_gem_v2 = exp_deaths_2020_year_gem %>% 
  dplyr::arrange(canton,GMDNR) %>% 
  dplyr::left_join(covs3)
```

```{r include=FALSE}
exp_deaths_2020_year_agg_v2 %<>% 
  filter(!is.na(vote_yes_jun_cat))

exp_deaths_2020_year_gem_v2 %<>% 
  filter(!is.na(vote_yes_jun_cat))

exp_deaths_2020_year_agg_v2 %<>% 
  filter(!is.na(vote_yes_nov_cat)) 

exp_deaths_2020_year_gem_v2 %<>% 
  filter(!is.na(vote_yes_nov_cat))

write_rds(exp_deaths_2020_year_v2, "results_v2/exp_deaths_2020_year.Rds")
write_rds(exp_deaths_2020_year_agg_v2, "results_v2/exp_deaths_2020_year_agg.Rds")
write_rds(exp_deaths_2020_year_gem_v2, "results_v2/exp_deaths_2020_year_gem.Rds")
```
